### 2023-02-22 doc/manuscript/translocation.qmd
Add fisher citation and adjust intro wording

^3^Department of Forestry, Wildlife, and Fisheries, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996

@@

Emerging infectious diseases are increasing [@daszak2000] and pose a severe threat to human, livestock, and wildlife populations. In wildlife populations, disease is responsible for dramatic declines, extirpations, and even extinctions in mammals, birds, amphibians, echinoderms, and other taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021]. Although the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting and understanding disease impacts [@langwig2015], this information is generally insufficient to allow the development of effective methods to recover disease-impacted species [@scheele2021; @fellers2007]. Importantly, some populations previously decimated by disease have subsequently recovered to varying degrees [@newell2013; @voyles2018; @knapp2016] and might provide critical insights for species conservation efforts. Unfortunately, given the nascent nature of wildlife recoveries in the presence of disease, we know far less about the mechanisms driving population recovery [@brannelly2021; @russell2020] and how these mechanisms can be leveraged to re-establish host populations across large spatial scales [@mendelson2019]. At a time when novel infectious diseases are emerging at an unprecedented rate, this knowledge gap has far-reaching consequences for global biodiversity.

@@

Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some frog populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016]. Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant genotypes [@knapp2016]. Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018]. Here, we report the results of a greatly expanded MYL frog reintroduction effort and an associated analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd. Collectively, the results show that MYL frogs generally meet the four conditions for reintroduction feasibility outlined above, and provide a novel example of how the reintroduction of resistant frogs can allow the large-scale recovery of this disease-impacted amphibian. These results may have broad implications for the many species worldwide that are threatened with extinction by emerging diseases.

@@

<!--# Let's revisit whether the selection results should go here or at the end of the results. Putting those results here would reduce the need to describe populations as "putatively resistant". -->

@@

To identify drivers of post-translocation frog survival, we developed a hierarchical model in a Bayesian framework (see SI for details). Our specific interest was whether Bd load had an important effect on adult survival, as would be expected if the effect of frog resistance was insufficient to overcome the negative effect of Bd. The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens). Important predictors included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996; @fig-cond-effects). Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects). In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens). In summary, our analyses show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved increased resistance and/or tolerance to Bd infection. In addition, the small changes in Bd load following translocation and relatively high survival in the majority of translocated cohorts suggests that putatively resistant frogs from recovering populations can maintain the benefits of resistance (i.e., low-moderate Bde loads, minimal Bd-induced mortality despite ongoing infection) in non-natal habitats. Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population that are the progeny of translocated individuals, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults <!--# Show results. -->. Collectively, these results show that putative resistance in non-natal habitats is maintained across generations of frogs.

@@

To test whether the observed yearly survival probabilities of putatively resistant adults in translocated populations was sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity. We parameterized the model using mark-recapture data and known life history values in this system (@tbl-param_values). Given observed yearly adult survival probabilities and a baseline yearly juvenile recruitment probability of 0.3, our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations). When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A). In contrast, for the other six translocated populations where yearly adult survival was less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A). A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).

Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of adult recruitment $\omega$, our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B). Therefore, even under conservative assumptions about recruitment probability $\omega$, these populations had some probability of persistence over 50 years following a single introduction. For these six populations, as $\omega$ increased from 0.1 to 0.3, 50 year extinction probability was less 0.5, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B). In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1. To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years). Overall, our model demonstrates that given observed yearly adult survival probabilities of putatively resistance frogs in non-natal locations, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.

>>>>

^3^School of Natural Resources, University of Tennessee Institute of Agriculture, Knoxville, TN, 37996

@@

Emerging infectious diseases are increasing [@daszak2000] and pose a severe threat to human, livestock, and wildlife populations. In wildlife populations, disease is responsible for dramatic declines, extirpations, and even extinctions in mammals, birds, amphibians, echinoderms, and other taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021]. Although the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting and understanding disease impacts [@langwig2015], this information is generally insufficient to allow the development of effective methods to recover disease-impacted species [@scheele2021; @fellers2007]. Importantly, some populations previously decimated by disease have subsequently recovered to varying degrees [@newell2013; @voyles2018; @knapp2016] and might provide critical insights for species conservation efforts. Unfortunately, given the nascent nature of wildlife recoveries in the presence of disease, we know far less about the mechanisms driving population recovery [@brannelly2021; @russell2020] and how these mechanisms can be leveraged to re-establish host populations across large spatial scales [@mendelson2019]. At a time when novel infectious diseases are emerging at an unprecedented rate [@fisher2012], this knowledge gap has far-reaching consequences for global biodiversity.

@@

Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some frog populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016]. Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant genotypes [@knapp2016]. Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018]. Here, we report the results of a greatly expanded MYL frog reintroduction effort, a long-term viability modeling, and an associated analysis to evaluate genomic signatures of selection following MYL frog exposure to Bd. Collectively, the results show that MYL frogs generally meet the four conditions for reintroduction feasibility outlined above, and provide a novel example of how the reintroduction of resistant frogs can allow the large-scale recovery of this disease-impacted amphibian. These results may have broad implications for the many species worldwide that are threatened with extinction by emerging diseases.

@@

<!--# Let's revisit whether the selection results should go here or at the end of the results. Putting those results here would reduce the need to describe populations as "putatively resistant". Mark W: Agreed. -->

@@

To identify drivers of post-translocation frog survival, we developed a hierarchical model in a Bayesian framework (see SI for details). Our specific interest was whether Bd load had an important effect on adult survival, as would be expected if the effect of frog resistance was insufficient to overcome the negative effect of Bd. The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens). Important predictors included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996; @fig-cond-effects). Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects). In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens). In summary, our analyses show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved increased resistance to Bd infection. In addition, the small changes in Bd load following translocation and relatively high survival in the majority of translocated cohorts suggests that putatively resistant frogs from recovering populations can maintain the benefits of resistance (i.e., low-moderate Bde loads, minimal Bd-induced mortality despite ongoing infection) in non-natal habitats. Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population that are the progeny of translocated individuals, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults <!--# Show results. -->. Collectively, these results show that putative resistance in non-natal habitats is maintained across generations of frogs.

@@

To test whether the observed yearly survival probabilities of putatively resistant adults in translocated populations was sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity. We parameterized the model using mark-recapture data and known life history values in this system (@tbl-param_values). Given observed yearly adult survival probabilities and a baseline yearly juvenile recruitment probability of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations). When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A). In contrast, for the other six translocated populations where yearly adult survival was less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A). A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).

Given demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to adults $\omega$, our model predicted that, following a single introduction of 40 adult individuals into a population, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of less than 1 (@fig-viability B). Therefore, even under conservative assumptions about recruitment probability $\omega$ (a parameter where ), these populations had some probability of persistence over 50 years following a single introduction. For these six populations, as $\omega$ increased from 0.1 to 0.3, 50 year extinction probability was less 0.5, indicating strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment success (@fig-viability B). In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1. To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years). Overall, our model demonstrates that given observed yearly adult survival probabilities of resistance frogs in non-natal locations, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
